Moisture effects on leaf litter are species-specific and result in non-additive flammability in mixed conifer forest

### 2019-07-25 ms/Magalhaes_Schwilk_litter_flam_moisture.tex
Format for Journal of ecology
Still some things to do:

1. Check journal limits on number of figures, move to SI if necessary
2. Check figures for font sizes, correct "dessication rate", etc.
3. Correct some bib isse (Nelson Jr, etc)
4. Check table formatting and spacing
5. Finnish and proof SI

%% reftex-default-bibliography: ("/home/schwilk/write/bib/schwilk.bib")

% LaTeX support: latex@mdpi.com 
%  In case you need support, please attach all files that are necessary for compiling as well as the log file, and specify the details of your LaTeX setup (which operating system and LaTeX version  tools you are using).

%  =================================================================
\documentclass[fire,article,submit,moreauthors,pdftex]{Definitions/mdpi}

%\usepackage[utf8]{inputenc}

% If you would like to post an early version of this manuscript as a preprint, you may use preprint as the journal and change 'submit' to 'accept'. The document class line would be, e.g., \documentclass[preprints,article,accept,moreauthors,pdftex]{mdpi}. This is especially recommended for submission to arXiv, where line numbers should be removed before posting. For preprints.org, the editorial staff will make this change immediately prior to posting.

%--------------------
% Class Options:
%--------------------

%----------
% journal
%----------
%=================================================================
\firstpage{1} 
\makeatletter 
\setcounter{page}{\@firstpage} 
\makeatother
\pubvolume{xx}
\issuenum{1}
\articlenumber{5}
\pubyear{2019}
\copyrightyear{2019}
%\externaleditor{Academic Editor: name}
\history{Received: date; Accepted: date; Published: date}
%\updates{yes} % If there is an update available, un-comment this line

%% MDPI internal command: uncomment if new journal that already uses continuous page numbers 
%\continuouspages{yes}

%------------------------------------------------------------------
% The following line should be uncommented if the LaTeX file is uploaded to arXiv.org
%\pdfoutput=1

%=================================================================
% Add packages and commands here. The following packages are loaded in our class file: fontenc, calc, indentfirst, fancyhdr, graphicx, lastpage, ifthen, lineno, float, amsmath, setspace, enumitem, mathpazo, booktabs, titlesec, etoolbox, amsthm, hyphenat, natbib, hyperref, footmisc, geometry, caption, url, mdframed, tabto, soul, multirow, microtype, tikz

%\bibliography{/home/schwilk/write/bib/schwilk.bib}

%=================================================================
%% Please use the following mathematics environments: Theorem, Lemma, Corollary, Proposition, Characterization, Property, Problem, Example, ExamplesandDefinitions, Hypothesis, Remark, Definition
%% For proofs, please use the proof environment (the amsthm package is loaded by the MDPI class).

%=================================================================
% Full title of the paper (Capitalized)
\Title{Moisture effects on leaf litter are species-specific and result in non-additive flammability in mixed conifer forest}

% Author Orchid ID: enter ID or remove command
\newcommand{\orcidauthorA}{0000-0000-000-000X} % Add \orcidA{} behind the author's name
%\newcommand{\orcidauthorB}{0000-0000-000-000X} % Add \orcidB{} behind the author's name

% Authors, for the paper (add full first names)
\Author{Rita Quinones de Magalhaes $^{\dagger}$ and Dylan W. Schwilk}

% Authors, for metadata in PDF
\AuthorNames{Rita Magalh\~{a}es, Dylan W. Schwilk}

% Affiliations / Addresses (Add [1] after \address if there is only one affiliation.)
\address[1]{Texas Tech University}

% Contact information of the corresponding author
\corres{Correspondence: dylan.schwilk@ttu.edu, DWS}

% Current address and/or shared authorship
\firstnote{Current address: Rochester Institute of Technology, Rochester, NY 14623} 
%\secondnote{These authors contributed equally to this work.}
% The commands \thirdnote{} till \eighthnote{} are available for further notes

%\simplesumm{} % Simple summary

%\conference{} % An extended version of a conference paper

% Abstract (Do not insert blank lines, i.e. \\) 
\abstract{ Moisture content is a strong determinant of forest fuel flammability
  and varies on small spatial scales and over short time periods. Emperical
  modeling of equilibrium dead fine fuel moisture is well established, but less
  is known about the mechanisms by which species-specific litter traits
  influence litter moisture dynamics, particularly the processes of moisture
  absorption and desorption. Characterizing how species’ leaf litter retain and
  release moisture through time is critical to the assessment of fuel’s
  availability to burn. Furthermore, Natural forests are often comprised of
  mixed stands, leading to mixtures of leaf litter on the surface fuelling the
  fires. Multiple studies have established the existence of non-additive
  effects in the flammability of dry litter, but we lack an understanding of
  moisture synergies in mixed litter and of moisture effects on the
  flammability of mixed litter. The litter from eight species of a
  mixed-conifer forest was saturated and allowed to dry to determine moisture
  absorption capacity and desorption rates. Burn trials were performed in moist
  litter beds of single species and of mixtures to establish flammability
  response to the dry-down process. Moisture dynamics vary across species with
  higher specific leaf area being associated with higher maximum absorption and
  lower litter bulk density being associated with faster drying rates. Species
  that produce more aerated litter beds (such as \emph{Q. kelloggii} and
  \emph{Pinus} species) have faster rates of litter drying. We examined two
  axes of flammability variation (spread rate and fuel consumption) and found
  that these differing moisture dynamics result in time since wetting having
  strongly differing effects on flammability across the major litter types
  typified by different tree genera. Finally, we found that litter mixtures
  exhibited consistent non-additivity in flame spread rate in fuel consumption.
  This non-additivity became more positive with fuel drying. Litter mixtures
  burn with fire behavior more similar to that of the most flammable
  constituent species and this effect increases as fuels dry. ADD Synthesis
  statement.}

% Keywords
\keyword{flammability, litter, flame spread rate}

%\setcounter{secnumdepth}{4}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{document}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Introduction}

Fuel moisture content is an important, and often driving, determinant of
surface fire behavior. Environmental conditions such as temperature, relative
humidity, and wind drive the moisture content of the fuel available to burn
\cite{Kreye+Varner+etal-2018}. In particular, fine dead fuels respond very
readily to environmental changes \cite{Nelson-2001}. Fuel moisture influences
flammability and fire behavior through effects on ignition, spread rate, flame
height, fuel consumption, and heat release, due to the high specific heat
content of water \cite{Rothermel-1972, Nelson-2011}. Moisture acts as a
heat-sink resulting in less energy available for propagation, and the resulting
water vapor will both cool the flaming front and dilute flammable gases
\cite{Albini-1976, Shafizadeh-1977}. Variation in moisture of living fuels has
strong effects on flammability, but moisture of dead excised litter fuels can
also vary. Moisture of litter fuels varies with climate, with seasons, and with
short term weather within a season. Therefore, the rate of drying and
flammability at any particular moisture content are key parameters that
influence fuel flammability throughout the fire season \cite{Nelson-2001}.

Fire scientists often categorize fuel by size classes that correspond to
moisture exchange rates, although such simplified categories have long been
recognized as not sufficient to describe species-specific effects on moisture
dynamics \cite{Anderson-1985}. Fuel moisture dynamics of leaf litter are
dependent upon forest structure (shading and rain throughfall effects), litter
physical structure, and chemical composition of the fuel-beds
\cite{Nelson+Hiers-2008, Matthews-2014, Kreye_Hiers_etal-2018}. A variety of
models describing moisture dynamics of fine fuels are used by fire scientists
\cite{Viney-1991, Nelson-2000, Catchpole+Catchpole+etal-2001}. These models
usually aim to describe equilibrium conditions and at least include temperature
and relative humidity although some include precipitation and solar radiation
effects as well. Explicit models of fire behavior as a function of time since
wetting could improve fire prediction, however. In many litter fuel driven
systems, natural lightning ignitions are concurrent with precipitation and,
therefore, fire behavior can depend on rate of drying following a wetting
event. If species differ in their drying rate, this may strongly influence
relatively flammability across species at any particular time following
wetting. Such differential moisture content across time, may contribute to the
often high spatial heterogeneity of surface fires
\cite{Knapp_Schwilk_etal-2006, Kreye_Hiers_etal-2018}.

% Fine fuel moisture content reduces ignition probability and flame spread rate \cite{Gisborne-1936,   Fons-1946, Anderson+Rothermal-1965}.

Litter packing (measured as litter-bed density or packing ratio) controls the
flammability of dry litter fuels in seasonally dry climates because higher
litter density reduces flammability, especially on the flame spread rate axis.
Leaf litter density is driven by leaf size and shape
\cite{Fonda+Belanger+etal-1998, Scarff+Westoby-2006, Kane+Varner+etal-2008}.
These litter particle effects on litter bed aeration, in turn, have effects on
litter bed flammability \cite{Ganteaume+Marielle+etal-2011,
  Schwilk+Caprio-2011, Kreye+Varner+etal-2013}. These species-specific effects
are well established, however, how moisture dynamics modulate species specific
flammability is poorly known, and could be driven by by leaf traits, emergent
litter litter traits, or both. Species specific effects on litter flammability
are recorded for the forests of the southeastern US \cite{Nowacki+Abrams-2008}
where species-specific litter drying rates play a role in flammability
feedbacks \cite{Kreye+Varner+etal-2013}. Similarly, in mixed-conifer forest of
the western USA, where the fire regime is characterized by surface fires fueled
by plant litter, overstory tree composition drives variation in local fire
behavior \cite{Schwilk+Caprio-2011} as a result of differential litter
flammability cross tree species and as a result of synergistic interactions in
multi-species litter mixtures \cite{Magalhaes+Schwilk-2012}. These flammability
feedbacks may help explain historical fire frequency patterns, but current
estimates of litter flammability in mixed conifer forests are based largely on
burning dry litter.

Fuel moisture content through time is controlled by two main processes,
absorptive capacity and rate of moisture loss \citep{Kreye+Kobziar+etal-2013}.
Absorptive capacity represents the potential maximum fuel moisture and is
evaluated by the moisture content of litter when saturated. After saturation,
the differences in moisture content indicate different abilities to retain
moisture. Litter particle geometry may have direct effects on maximum moisture
absorption through area-mass relationships and we also expect effects of leaf
morphology, especially leaf cuticle thickness \cite{Van_Wagner-1969}. Drying
rates should also be influenced by leaf and litter particle traits:
surface-area-to-volume ratio and litter bulk density will influence air flow
through the litter bed. Litter packing therefore has two pathways of effect on
fire behavior: directly by its effect on oxygen availability for combustion,
and indirectly by its effect on dry-down rate, and both these pathways are
affected by leaf litter traits, especially leaf size
\cite{Scarff+Westoby-2006}. Fuel-beds composed of larger leaves with lower
moisture retainability and should favor drying, thus increasing the probability
of ignition and fire spread.

Flammability is a multi-dimensional characteristic of fuels
\citep{Schwilk-2015, Pausas+Keeley+etal-2017} with different traits influencing
different components of fire behavior. In litter driven fire, the two major
axes of variation are described by measurements related to flame spread rate
and measurements related to heat transfer. Flame spread rate is easy to measure
in burning trials, but heat transfer can be more difficult and often proxies
such as duration of heating are used \citep{Magalhaes+Schwilk-2012,
  Varner+Kane+etal-2015}. These different components of fire behavior have
differing ecological effects. Species traits may interact such that the
flammability of multi-species mixtures is not the average of the constituent
species. Such non-additive effects have been shown to be synergistic in dry
litter fuels where the flammability of a mixture is driven by the most
flammable species \cite{VanAltena+Logtestjin+etal-2012,
  Magalhaes+Schwilk-2012}. It is possible that similar non-additive effects may
influence litter moisture content, especially if different leaf traits
influence absorptive capacity and drying rate. There is some evidence for
non-additive flammability in moist fuels with one study finding increasing
non-additivity with moisture \cite{Blauw+Wensink+etal-2015}.

We investigated species specific litter moisture dynamics across eight tree
species in a temperate fire-prone forest in California. We determined the
maximum water retention and rate of moisture loss and examined how these vary
in relation to two leaf and litter bed traits. We then examined the influence
of moisture on the flammability across these species. Experimental burn trials
of leaf litter at various moisture levels determined how time since wetting
influenced two axes of flammability characterized by flame spread rate and
total fuel consumed. Previous work described the flammability of these species
in monocultures and in mixtures for oven-dry litter
\cite{Magalhaes+Schwilk-2012} and found consistent positive non-additivity of
flammability where fire behavior of mixtures was driven by the most flammable
species. Here, we examined if this positive non additivity was maintained in
mixtures across varying levels of fuel moisture. The burn trials were also
performed with multi-species litter mixtures to evaluate if species contribute
equally to litter moisture dynamics or if there are non-additive effects as
observed in oven-dried litter. We tested if the non-additive effects on
reported in previous studies occur under ecologically-relevant moisture levels
and whether fuel moisture effects strengthen or mask such non-additivity.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Materials and Methods}

\subsection{Site description and species selection}

@@

\subsection{Moisture absorption and desorption}

@@

With only eight species in this study, our power to test leaf trait effects on
the observed moisture dynamics was low, so we decided \emph{a priori} to test
the effect of one leaf trait likely to influence moisture absorption, specific
leaf area, and one litter trait known to directly effect flammability and
likely to influence drying rate, litter bulk density. Specific leaf area was
taken from leaf fresh leaf area and leaf dry mass from 5 individuals per
species and was collected as part of a previous study
\cite{Magalhaes+Schwilk-2012}. Litter bulk density was obtained in this
experiment for each basket by dividing the dry mass (450\,g) by the estimated
litter volume (basket width x length x litter depth).

\subsection{Flammability assessment}

We performed flammability tests across multiple moisture levels on monospecific
litter beds and on mixed litter beds with each species or mixture replicated 5
times at each moisture level. We used the initial dry-down experiment to guide
our timing so as to attempt to burn samples over a range of moisture contents.
We aimed for moisture contents from 10\% to 100\% on a dry mass basis as
preliminary trials on damp fuels suggested that across these fuels ignition is
unlikely above 100\% fuel moisture. To expand our data on the dry end of this
gradient, we supplemented these data with data from an earlier experiment on
oven-dried litter \cite{Magalhaes+Schwilk-2012}. This supplemental data added
five replicates per species at an oven dried moisture content of $<$ 5\%. We
assigned these replicates a time since wetting value of 144 hours equivalent to
the longest time since drying for the wetted samples. Flammability was assessed
using a 150\,cm long burn table, 15\,cm wide and 15\,cm tall, in which leaf
litter was placed and the table gently shaken to allow settling. Burning trials
all used 450\,g (dry weight) of litter. For three-species mixtures, this was
150\,g per species. This design allows fire to reach constant flame spread
rates \cite{Magalhaes+Schwilk-2012}. Two graduated metal rulers equally spaced
along the apparatus allowed visual determination of maximum flame height.
Litter was ignited with a propane torch; maximum adiabatic flame temperature of
1899 C, \url{http://www.benzomatic.com/products/fuel.aspx}) at one end of the
apparatus and allowed to burn until extinction. If a sample failed to ignite
after 30 seconds, no further attempt was made and if three samples of a species
at a particular moisture level failed to ignite, no further attempts were made
on that species/moisture combination and the replicate was allowed to continue
drying. A timer was used to record total time to ignition and duration of
flaming combustion. The flammability trials followed the methodology detailed
in de Magalhães and Schwilk \cite{Magalhaes+Schwilk-2012}.

During flammability trials, we recorded time to ignition (s), duration of
flaming combustion (s), flame spread rate (mm/s), calculated by dividing the
length of the burned surface by the time it takes the flaming front to reach
the end of the apparatus, maximal flame height (mm), and fuel consumption (\%).
Some trials resulted only in smoldering combustion; for those trials, a time to
ignition and mass loss were recorded, but flame spread rate and flame height
were recorded as zero. Our thermocouple system failed or data was lost for two
days of burning trials, therefore we have omitted those data from this
analysis. Our past work \cite{Magalhaes+Schwilk-2012} demonstrated that flame
spread rate captures one important axis of flammability variation
\citep{Schwilk-2015, Pausas+Keeley+etal-2017} and percent fuel consumed
captures a portion of the second major axis of flammability variation (although
not as well as integrating temperature over time does). Our analysis,
therefore, focused on these two relatively orthogonal measures of flammability
to avoid examining multiple redundant dependent variables.

@@

\subsection{Statistical analyses}

@@

Where $m$ is moisture content at time $t$, and $\lambda$ is the exponential
decay coefficient. To estimate these coefficients we fit a linear mixed-effects
model by first taking the natural logarithm of moisture content, with time
since wetting and species as fixed effects nested within replicate basket. We
extracted the coefficients of the fitted model to describe maximum water
retainability ($m_{max}$; g of water per g of dry mass) and desiccation rate
($\lambda$; hour$^{-1}$) for each species. All analyses were conducted in R
\cite{RCoreTeam-2019}. We fit models with R packages `lme4'
\cite{Bates_Machler_etal-2015} and `afex' \cite{Singmann_Bolker_etal-2017}.
Approximate degrees of freedom and p-values were calculated by the
Kenward-Roger approximation \cite{Kenward_Roger-1997} which is recommended by
Luke \cite{Luke-2017} as producing acceptable Type 1 error rates. We tested if
one leaf trait (specific leaf area) and one litter trait (litter bulk density)
influenced moisture dynamics by fitting linear models with species means as
observations. For each response variable we fit a single linear model with two
predictors and an interaction term.

Our past work \cite{Magalhaes+Schwilk-2012} demonstrated that dry litter
flammability across these eight species was driven by leaf traits and litter
packing. These previously reported flammability trials demonstrated that
species within a genus tended to behave similarly and the two scale-leaved
genera in the Cupressaceae, \emph{Calocedrus} and \emph{Sequoiadendron}, also
behaved similarly to one another. Therefore, to preserve power used taxonomic
groups (\emph{Abies}, Cupressaceae, \emph{Quercus}, and \emph{Pinus}) as our
flammability functional groups and treated these as fixed effects in our
models. The effect of fuel moisture and time since wetting on flammability
parameters were modeled using linear models with fuel moisture or time since
wetting and taxon as fixed effects and average vapor pressure deficit in the
burning room during that trial as a nuisance covariate. We used linear mixed
models to predict spread rate, but fuel consumption tends to exhibit threshold
effects with few intermediate values. Therefore, we used mixed generalized
linear models with a beta regression link with the `glmmADMB` package
\cite{Skaug_Fournier_etal-2016} to model percent fuel consumed as a function
moisture content and of time since wetting.

We investigated possible non-additive effects on moisture and on flammability
by testing if moisture content and flammability of mixtures were predicted by
the average of the constituent species. To choose mixtures, we first grouped
the species according to their maximum water retainability and dessication rate
according to estimated marginal means by species using the R package `emmeans'
\cite{Lenth-2019}. This resulted in four groups of species and we then selected
one representative species from each litter type group and produced all four
possible three-species litter mixtures using those four species. For each
mixture and time since wetting, we predicted moisture content based on the mean
content of the three constituent species at that time since wetting. We
predicted flammability parameters based on the average flammability measures
from individual species trials at each time since wetting and then calculated a
predicted spread rate and fuel consumption as the average of these three values
for the mixture. The expectation from the null model was that the difference
between observed and predicted values of flammability will be zero. We tested
if there were significant departures from the null (zero) with a mixed effects
linear model. For flammability parameters, we fit linear mixed effect models
with the observed-predicted flammability value as the dependent variable and
used Wald tests to test if the centered intercept was different than zero (a
measure of overall non-additivity) and if the slope with time since wetting was
different from zero (indicating non-additivity changing with drying).

>>>>

\documentclass[letterpaper,12pt]{article}
\usepackage{jae/jae}
\usepackage{booktabs}

\title{Moisture effects on leaf litter are species-specific and result in non-additive flammability in mixed conifer forest}

\running{Leaf litter flammability}

\author{Rita Quinones de Magalhaes$^{1,2}$ and Dylan W. Schwilk$^1$}

\affiliations{
    \item Texas Tech University, Lubbock, Texas 79411, USA.
    \item Current address: Rochester Institute of Technology, Rochester, NY 14623, USA.
}

\nwords{X}
\ntables{X}
\nfig{X}
\nref{X}

\corr{\url{dylan.schwilk@ttu.edu}}

\begin{document}

\maketitle

\begin{abstract}
  \noindent \begin{enumerate}

  \item Moisture content is a strong determinant of forest fuel flammability
  and varies on small spatial scales and over short time periods. Empirical
  modeling of equilibrium dead fine fuel moisture is well established, but less
  is known about the mechanisms by which species-specific litter traits
  influence litter moisture dynamics, particularly the processes of moisture
  absorption and desorption. Characterizing how species’ leaf litter retain and
  release moisture through time is critical to the assessment of fuel’s
  availability to burn.
  \item Natural forests are often comprised of
  mixed stands, leading to mixtures of leaf litter on the surface fuelling the
  fires. Multiple studies have established the existence of non-additive
  effects in the flammability of dry litter, but we lack an understanding of
  moisture synergies in mixed litter and of moisture effects on the
  flammability of mixed litter.
  \item The litter from eight species of a
  mixed-conifer forest was saturated and allowed to dry to determine moisture
  absorption capacity and desorption rates. Burn trials were performed in moist
  litter beds of single species and of mixtures to establish flammability
  response to the dry-down process.
  \item Moisture dynamics varied across species with
  higher specific leaf area being associated with higher maximum absorption and
  lower litter bulk density being associated with faster drying rates. Species
  that produce more aerated litter beds (such as \emph{Q. kelloggii} and
  \emph{Pinus} species) have faster rates of litter drying.
  \item We examined two
  axes of flammability variation (spread rate and fuel consumption) and found
  that these differing moisture dynamics result in time since wetting having
  strongly differing effects on flammability across the major litter types
  typified by different tree genera. We found that litter mixtures
  exhibited consistent non-additivity in flame spread rate in fuel consumption.
  This non-additivity became more positive with fuel drying.
    \item \emph{Synthesis:} Litter mixtures burn with fire
  behavior more similar to that of the most flammable constituent species and
  this effect increases as fuels dry. statement. Such positive feedback effects
  increases the potential for species to exert community scale effects even
  when relatively rare and that can influence community assembly and the
  evolutionary of niche-constructing traits.
  \end{enumerate}
\end{abstract}

\noindent \textbf{Keywords:} California, flame spread rate, flammability, fuel consumption, fuel moisture, leaf litter, non-additivity

\newpage

\section*{Introduction}

Fuel moisture content is an important, and often driving, determinant of
surface fire behavior. Environmental conditions such as temperature, relative
humidity, and wind drive the moisture content of the fuel available to burn
\citep{Kreye+Varner+etal-2018}. In particular, fine dead fuels respond very
readily to environmental changes \citep{Nelson-2001}. Fuel moisture influences
flammability and fire behavior through effects on ignition, spread rate, flame
height, fuel consumption, and heat release, due to the high specific heat
content of water \citep{Rothermel-1972, Nelson-2011}. Moisture acts as a
heat-sink resulting in less energy available for propagation, and the resulting
water vapor will both cool the flaming front and dilute flammable gases
\citep{Albini-1976, Shafizadeh-1977}. Variation in moisture of living fuels has
strong effects on flammability, but moisture of dead excised litter fuels can
also vary. Moisture of litter fuels varies with climate, with seasons, and with
short term weather within a season. Therefore, the rate of drying and
flammability at any particular moisture content are key parameters that
influence fuel flammability throughout the fire season \citep{Nelson-2001}.

Fire scientists often categorize fuel by size classes that correspond to
moisture exchange rates, although such simplified categories have long been
recognized as not sufficient to describe species-specific effects on moisture
dynamics \citep{Anderson-1985}. Fuel moisture dynamics of leaf litter are
dependent upon forest structure (shading and rain throughfall effects), litter
physical structure, and chemical composition of the fuel-beds
\citep{Nelson+Hiers-2008, Matthews-2014, Kreye_Hiers_etal-2018}. A variety of
models describing moisture dynamics of fine fuels are used by fire scientists
\citep{Viney-1991, Nelson-2000, Catchpole+Catchpole+etal-2001}. These models
usually aim to describe equilibrium conditions and at least include temperature
and relative humidity although some include precipitation and solar radiation
effects as well. Explicit models of fire behavior as a function of time since
wetting could improve fire prediction, however. In many litter fuel driven
systems, natural lightning ignitions are concurrent with precipitation and,
therefore, fire behavior can depend on rate of drying following a wetting
event. If species differ in their drying rate, this may strongly influence
relatively flammability across species at any particular time following
wetting. Such differential moisture content across time, may contribute to the
often high spatial heterogeneity of surface fires
\citep{Knapp_Schwilk_etal-2006, Kreye_Hiers_etal-2018}.

% Fine fuel moisture content reduces ignition probability and flame spread rate \citep{Gisborne-1936,   Fons-1946, Anderson+Rothermal-1965}.

Litter packing (measured as litter-bed density or packing ratio) controls the
flammability of dry litter fuels in seasonally dry climates because higher
litter density reduces flammability, especially on the flame spread rate axis.
Leaf litter density is driven by leaf size and shape
\citep{Fonda+Belanger+etal-1998, Scarff+Westoby-2006, Kane+Varner+etal-2008}.
These litter particle effects on litter bed aeration, in turn, have effects on
litter bed flammability \citep{Ganteaume+Marielle+etal-2011,
  Schwilk+Caprio-2011, Kreye+Varner+etal-2013}. These species-specific effects
are well established, however, how moisture dynamics modulate species specific
flammability is poorly known, and could be driven by by leaf traits, emergent
litter litter traits, or both. Species specific effects on litter flammability
are recorded for the forests of the southeastern US \citep{Nowacki+Abrams-2008}
where species-specific litter drying rates play a role in flammability
feedbacks \citep{Kreye+Varner+etal-2013}. Similarly, in mixed-conifer forest of
the western USA, where the fire regime is characterized by surface fires fueled
by plant litter, overstory tree composition drives variation in local fire
behavior \citep{Schwilk+Caprio-2011} as a result of differential litter
flammability cross tree species and as a result of synergistic interactions in
multi-species litter mixtures \citep{Magalhaes+Schwilk-2012}. These flammability
feedbacks may help explain historical fire frequency patterns, but current
estimates of litter flammability in mixed conifer forests are based largely on
burning dry litter.

Fuel moisture content through time is controlled by two main processes,
absorptive capacity and rate of moisture loss \citep{Kreye+Kobziar+etal-2013}.
Absorptive capacity represents the potential maximum fuel moisture and is
evaluated by the moisture content of litter when saturated. After saturation,
the differences in moisture content indicate different abilities to retain
moisture. Litter particle geometry may have direct effects on maximum moisture
absorption through area-mass relationships and we also expect effects of leaf
morphology, especially leaf cuticle thickness \citep{Van_Wagner-1969}. Drying
rates should also be influenced by leaf and litter particle traits:
surface-area-to-volume ratio and litter bulk density will influence air flow
through the litter bed. Litter packing therefore has two pathways of effect on
fire behavior: directly by its effect on oxygen availability for combustion,
and indirectly by its effect on dry-down rate, and both these pathways are
affected by leaf litter traits, especially leaf size
\citep{Scarff+Westoby-2006}. Fuel-beds composed of larger leaves with lower
moisture retainability and should favor drying, thus increasing the probability
of ignition and fire spread.

Flammability is a multi-dimensional characteristic of fuels
\citep{Schwilk-2015, Pausas+Keeley+etal-2017} with different traits influencing
different components of fire behavior. In litter driven fire, the two major
axes of variation are described by measurements related to flame spread rate
and measurements related to heat transfer. Flame spread rate is easy to measure
in burning trials, but heat transfer can be more difficult and often proxies
such as duration of heating are used \citep{Magalhaes+Schwilk-2012,
  Varner+Kane+etal-2015}. These different components of fire behavior have
differing ecological effects. Species traits may interact such that the
flammability of multi-species mixtures is not the average of the constituent
species. Such non-additive effects have been shown to be synergistic in dry
litter fuels where the flammability of a mixture is driven by the most
flammable species \citep{VanAltena+Logtestjin+etal-2012,
  Magalhaes+Schwilk-2012}. It is possible that similar non-additive effects may
influence litter moisture content, especially if different leaf traits
influence absorptive capacity and drying rate. There is some evidence for
non-additive flammability in moist fuels with one study finding increasing
non-additivity with moisture \citep{Blauw+Wensink+etal-2015}.

We investigated species specific litter moisture dynamics across eight tree
species in a temperate fire-prone forest in California. We determined the
maximum water retention and rate of moisture loss and examined how these vary
in relation to two leaf and litter bed traits. We then examined the influence
of moisture on the flammability across these species. Experimental burn trials
of leaf litter at various moisture levels determined how time since wetting
influenced two axes of flammability characterized by flame spread rate and
total fuel consumed. Previous work described the flammability of these species
in monocultures and in mixtures for oven-dry litter
\citep{Magalhaes+Schwilk-2012} and found consistent positive non-additivity of
flammability where fire behavior of mixtures was driven by the most flammable
species. Here, we examined if this positive non additivity was maintained in
mixtures across varying levels of fuel moisture. The burn trials were also
performed with multi-species litter mixtures to evaluate if species contribute
equally to litter moisture dynamics or if there are non-additive effects as
observed in oven-dried litter. We tested if the non-additive effects on
reported in previous studies occur under ecologically-relevant moisture levels
and whether fuel moisture effects strengthen or mask such non-additivity.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section*{Materials and Methods}

\subsection*{Site description and species selection}

@@

\subsection*{Moisture absorption and desorption}

@@

With only eight species in this study, our power to test leaf trait effects on
the observed moisture dynamics was low, so we decided \emph{a priori} to test
the effect of one leaf trait likely to influence moisture absorption, specific
leaf area, and one litter trait known to directly effect flammability and
likely to influence drying rate, litter bulk density. Specific leaf area was
taken from leaf fresh leaf area and leaf dry mass from 5 individuals per
species and was collected as part of a previous study
\citep{Magalhaes+Schwilk-2012}. Litter bulk density was obtained in this
experiment for each basket by dividing the dry mass (450\,g) by the estimated
litter volume (basket width x length x litter depth).

\subsection*{Flammability assessment}

We performed flammability tests across multiple moisture levels on monospecific
litter beds and on mixed litter beds with each species or mixture replicated 5
times at each moisture level. We used the initial dry-down experiment to guide
our timing so as to attempt to burn samples over a range of moisture contents.
We aimed for moisture contents from 10\% to 100\% on a dry mass basis as
preliminary trials on damp fuels suggested that across these fuels ignition is
unlikely above 100\% fuel moisture. To expand our data on the dry end of this
gradient, we supplemented these data with data from an earlier experiment on
oven-dried litter \citep{Magalhaes+Schwilk-2012}. This supplemental data added
five replicates per species at an oven dried moisture content of $<$ 5\%. We
assigned these replicates a time since wetting value of 144 hours equivalent to
the longest time since drying for the wetted samples. Flammability was assessed
using a 150\,cm long burn table, 15\,cm wide and 15\,cm tall, in which leaf
litter was placed and the table gently shaken to allow settling. Burning trials
all used 450\,g (dry weight) of litter. For three-species mixtures, this was
150\,g per species. This design allows fire to reach constant flame spread
rates \citep{Magalhaes+Schwilk-2012}. Two graduated metal rulers equally spaced
along the apparatus allowed visual determination of maximum flame height.
Litter was ignited with a propane torch; maximum adiabatic flame temperature of
1899 C, \url{http://www.benzomatic.com/products/fuel.aspx}) at one end of the
apparatus and allowed to burn until extinction. If a sample failed to ignite
after 30 seconds, no further attempt was made and if three samples of a species
at a particular moisture level failed to ignite, no further attempts were made
on that species/moisture combination and the replicate was allowed to continue
drying. A timer was used to record total time to ignition and duration of
flaming combustion. The flammability trials followed the methodology detailed
in de Magalhães and Schwilk \citep{Magalhaes+Schwilk-2012}.

During flammability trials, we recorded time to ignition (s), duration of
flaming combustion (s), flame spread rate (mm/s), calculated by dividing the
length of the burned surface by the time it takes the flaming front to reach
the end of the apparatus, maximal flame height (mm), and fuel consumption (\%).
Some trials resulted only in smoldering combustion; for those trials, a time to
ignition and mass loss were recorded, but flame spread rate and flame height
were recorded as zero. Our thermocouple system failed or data was lost for two
days of burning trials, therefore we have omitted those data from this
analysis. Our past work \citep{Magalhaes+Schwilk-2012} demonstrated that flame
spread rate captures one important axis of flammability variation
\citep{Schwilk-2015, Pausas+Keeley+etal-2017} and percent fuel consumed
captures a portion of the second major axis of flammability variation (although
not as well as integrating temperature over time does). Our analysis,
therefore, focused on these two relatively orthogonal measures of flammability
to avoid examining multiple redundant dependent variables.

@@

\subsection*{Statistical analyses}

@@

Where $m$ is moisture content at time $t$, and $\lambda$ is the exponential
decay coefficient. To estimate these coefficients we fit a linear mixed-effects
model by first taking the natural logarithm of moisture content, with time
since wetting and species as fixed effects nested within replicate basket. We
extracted the coefficients of the fitted model to describe maximum water
retainability ($m_{max}$; g of water per g of dry mass) and desiccation rate
($\lambda$; hour$^{-1}$) for each species. All analyses were conducted in R
\citep{RCoreTeam-2019}. We fit models with R packages `lme4'
\citep{Bates_Machler_etal-2015} and `afex' \citep{Singmann_Bolker_etal-2017}.
Approximate degrees of freedom and p-values were calculated by the
Kenward-Roger approximation \citep{Kenward_Roger-1997} which is recommended by
Luke \citep{Luke-2017} as producing acceptable Type 1 error rates. We tested if
one leaf trait (specific leaf area) and one litter trait (litter bulk density)
influenced moisture dynamics by fitting linear models with species means as
observations. For each response variable we fit a single linear model with two
predictors and an interaction term.

Our past work \citep{Magalhaes+Schwilk-2012} demonstrated that dry litter
flammability across these eight species was driven by leaf traits and litter
packing. These previously reported flammability trials demonstrated that
species within a genus tended to behave similarly and the two scale-leaved
genera in the Cupressaceae, \emph{Calocedrus} and \emph{Sequoiadendron}, also
behaved similarly to one another. Therefore, to preserve power used taxonomic
groups (\emph{Abies}, Cupressaceae, \emph{Quercus}, and \emph{Pinus}) as our
flammability functional groups and treated these as fixed effects in our
models. The effect of fuel moisture and time since wetting on flammability
parameters were modeled using linear models with fuel moisture or time since
wetting and taxon as fixed effects and average vapor pressure deficit in the
burning room during that trial as a nuisance covariate. We used linear mixed
models to predict spread rate, but fuel consumption tends to exhibit threshold
effects with few intermediate values. Therefore, we used mixed generalized
linear models with a beta regression link with the `glmmADMB` package
\citep{Skaug_Fournier_etal-2016} to model percent fuel consumed as a function
moisture content and of time since wetting.

We investigated possible non-additive effects on moisture and on flammability
by testing if moisture content and flammability of mixtures were predicted by
the average of the constituent species. To choose mixtures, we first grouped
the species according to their maximum water retainability and dessication rate
according to estimated marginal means by species using the R package `emmeans'
\citep{Lenth-2019}. This resulted in four groups of species and we then selected
one representative species from each litter type group and produced all four
possible three-species litter mixtures using those four species. For each
mixture and time since wetting, we predicted moisture content based on the mean
content of the three constituent species at that time since wetting. We
predicted flammability parameters based on the average flammability measures
from individual species trials at each time since wetting and then calculated a
predicted spread rate and fuel consumption as the average of these three values
for the mixture. The expectation from the null model was that the difference
between observed and predicted values of flammability will be zero. We tested
if there were significant departures from the null (zero) with a mixed effects
linear model. For flammability parameters, we fit linear mixed effect models
with the observed-predicted flammability value as the dependent variable and
used Wald tests to test if the centered intercept was different than zero (a
measure of overall non-additivity) and if the slope with time since wetting was
different from zero (indicating non-additivity changing with drying).
